% chktex-file 44
\section{Errata and Known Issues}

This section tracks the current limitations and known deviations from the full Ethernet specification for the \textbf{Mac} core.

\subsection{Errata}
There are currently no known functional bugs (errata) that deviate from the intended design of the 10GbE 64-bit architecture.

\subsection{Known Issues}

The following items are known limitations in the current release of the \textbf{Mac} core:

\begin{itemize}[noitemsep]
    \item \textbf{Strict 64-bit Architecture:} 
    The \textbf{Mac} core is strictly architected for 10GbE 64-bit data paths. While the \textit{dataW} parameter is present in the configuration class, the internal submodules \textit{Xgmii2Axis64} and \textit{Axis2Xgmii64} are strictly architected for a 64-bit data path.
    The \textit{dataW} parameter is guarded by a Scala \texttt{require} statement. Attempting to elaborate the design with \textit{dataW} $\neq$ 64 will result in a elaboration-time error.
    The parameter is retained in the configuration for potential future support of different Ethernet data widths, but currently only 64-bit operation is supported.
    
    \item \textbf{Control Frame Logic Bypassed:} 
    The optional MAC Control Frame (MCF) insertion and extraction logic is currently bypassed. 
    \begin{itemize}
        \item The \textbf{sAxisTx} and \textbf{mAxisRx} user interfaces are connected directly to the internal \textit{Axis2Xgmii64} and \textit{Xgmii2Axis64} submodules via an \textit{AxisTie} bridge.
        \item As a result, Link Flow Control (LFC) and Priority Flow Control (PFC) requests are not processed, and pause frames are neither generated nor interpreted.
        \item All control-related status outputs are tied to logic \textbf{'0'} to ensure stability in the statistics aggregator.
    \end{itemize}

    \item \textbf{Inactive Configuration Ports:} 
    Configuration inputs related to Flow Control (prefixed with \textit{cfgMcf}, \textit{cfgTxLfc}, and \textit{cfgTxPfc}) are currently disconnected from the data path and have no effect on frame processing.
\end{itemize}